Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-04-18 doc/manuscript.Rmd
Most of Melissas edits at this point. A couple more to go though.

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Previous work has relied on studying bacteria and phages in isolation, reducing them to two separate communities. These approaches have not captured how these communities interact to share information (e.g. horizontal gene transfer) and maintain stability. We developed and implemented a network-based analytical approach to provide an initial understanding of phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages infected which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate the diversity of phage-bacteria connectedness across the human body. We focused on establishing a foundational understanding of microbiome network structures across the human body. We found that gut and skin network structure was person-specific, and was not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. Altogether this study characterizes the diversity of microbiome networks across the human body, and provides a baseline for future studies to investigate the role of ecological networks in disease states.

**Word Count**: 236 / 250

@@

Viruses and bacteria are critical components to the human microbiome and play important roles in health and disease. Bacterial communities have been associated with diseases including a wide range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases including inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz], and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered viromes (virus communities consisting primarily of bacteriophages in humans) have also been associated with various diseases and environmental perturbations including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], spread of antibiotic resistance [@Modi:2013fia; @Hannigan:2015fz], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. These shifts in virus community composition are not reflections of the bacterial communities, but rather act in concert with bacteria as one single overall community [@Norman:2015kb; @Haerter:2014ii]. Individual phages within these communities are capable of lysing their bacterial hosts and modulating their functionality through horizontal gene transfer (i.e. transduction) and altering bacterial host gene expression.

Previous work has relied on studying bacterial and phage communities in isolation, reducing them to two separate communities. In reality, bacteria and phage communities are dynamic and complex. They frequently share genetic information and work together to maintain stable ecosystems. Removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. The conventional approach to understanding the human microbiome has been unable to capture how these communities interact and influence each other, but relationship-based network approaches would allow us to capture this information. We aimed to bridge this knowledge gap by characterizing human bacterial and phage communities by their relationships, leveraging machine learning and graph theory techniques. We focused on characterizing the network diversity of the human skin and gut microbiomes to provide a foundation for further studies of disease network dynamics and to gain broader insights into human microbiome network diversity across different body sites.

We first investigated human microbiome phage and bacteria networks by utilizing three published microbiome datasets with paired virus and bacterial metagenomic sequence sets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Moebus:1981kp; @Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models by its inclusion of protein interaction data, negative as well as positive interactions, and the use of a more sophisticated machine learning algorithm instead of simple linear models [@Edwards:2015iz].

Through this approach, we were able to go beyond contemporary isolated methods to provide a basic understanding of the community network dynamics associated with healthy human phages and bacteria. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. By understanding the healthy state of network structures across the human body, we will empower future studies to begin investigating how these community structures change in disease and their overall impact on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in microbiome phage-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were all subjected to virus-like particle (VLP) purification to eliminate other organismal DNA and allow for direct assessment of the actively replicating virome.

The publications we utilized were published over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)**. The bacterial and viral sequences from those publications were quality filtered and assembled into contigs. Because contig assembly most commonly returns genome fragments, we clustered related bacteria and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm. We interpreted these clusters as operational units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht]. The resulting contigs and OGUs demonstrated high sequence coverage and length **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**.

## Establishing a Model of Phage-Bacteria Interactions
We predicted which phage OGUs infected which bacteria using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. This training set contained diverse bacteria and phages, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because this was a DNA sequencing study, RNA phages were not considered in the analysis **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (phage infects a bacterium) and negative relationships (phage does not infect a bacterium). This built on previous work, which focused only on positive relationships, by allowing us to validate the false positive *and* false negative rates associated with our candidate models [@Edwards:2015iz].

Four phage and bacterial genomic markers were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families known to be associated with known protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed well, with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were minimally important. Approximately one third of the relationships yielded no score and were therefore unknown, unable to be assigned an interaction **(Figure \ref{RocCurve} C)**.

Next we used our random forest model to classify the relationships between bacteria and phages in the experimental datasets. The relationships within the three studies were used to construct one master network, containing the three study sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and abundance within the community (based on sequence count) were also stored in the multi-study master network to allow for effective parsing and downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The resulting master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` of which represented phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` that represented bacteria. Although the network was highly connected, not all relationships were present in all samples. Furthermore, relationships were weighted by relative abundance of their associated bacteria and phages, meaning lowly abundant relationships could be present but insignificant compared to those that were highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included almost all of the master network nodes and edges **(Figure \ref{RocCurve} E - F)**. The gut diet and twin sample sets each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, which indicated more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

## Individuality of Microbial Networks
Skin and gut community membership and diversity are highly personal, with people remaining more similar to themselves over time compared to other people [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. We determined whether this personal conservation extended to network structure using our microbiome network model. To accomplish this, we calculated the degree of dissimilarity between each subject's network based on phage and bacteria abundance, as well as centrality. Weighted eigenvector centrality was calculated for all bacteria and phages within each sample graph. This metric defines central phages as those that are highly abundant and infect many distinct bacteria which themselves are abundant and infected by many other phages. Similarly, bacterial centrality was defined as those bacteria that were both abundant and susceptible to numerous phages which infected many bacteria. We then calculated the similarity of community networks using the weighted eigenvector centrality of all nodes between all samples. Samples with similar network structures were interpreted as having similar capacities for maintaining stability and transmitting genetic material.

We used this network dissimilarity metric to test whether microbiome network structures were more similar within people over time compared to other people. We found that gut microbiome network structure clustered by person (ANOSIM p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, R = `r signif(interstats[c(interstats$site %in% "AS"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} A**). We found evidence that network dissimilarity within people over the 8-10 day sampling period was less than the average dissimilarity between that person and others, although this was not statistically significant (p-value = `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} B**). The lack of statistical confidence was likely due to the small sample size of this dataset. Although there was evidence for gut network conservation among individuals, we found no evidence for conservation of gut network structures within families. The gut network structures were not more similar between twins and their mothers (intrafamily) compared to other families of twins and mothers (inter-family) (p-value = `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} C**).

Skin microbiome network structure was strongly conserved within individuals, and more adequately confirmed by our larger skin dataset (p-value = `r signif(interstats[c(interstats$site %in% "Skin"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} D**). This distribution was similar when separated by anatomical sites, with most sites also being significantly more conserved within individuals **(Supplemental Figure \ref{allskin})**.

## Role of Diet & Obesity in Gut Microbiome Conectivity
Diet is a major environmental factor that influences resource availability and gut microbiome composition and diversity, including bacteria and phages [@Minot:2011ez; @Turnbaugh:2009hf; @David:2014cl]. Previous work in isolated culture-based systems has suggested that changes in nutrient availability are associated with altered phage - bacteria network structures, although this has yet to be tested in humans [@Poisot:2011jc]. We hypothesized that a change in diet would be associated with a change in network structure.

We evaluated the diet-associated differences in gut network structure by quantifying how central each sample's network was on average. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. Degree centrality, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. We more completely quantified the network properties by supplementing degree centrality with measurements of closeness centrality. Closeness centrality is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that genetic information or the effects of altered abundance would be more impactful to all other microbes in the system. A network with higher average closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience to instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

We found that gut microbiome network structures associated with high-fat diets were less connected than those of low-fat diets **(Figure \ref{dietnetworks} A-B)**. Tests for statistical differences were not performed due to the small sample size. High-fat diets exhibited less degree centrality **(Figure \ref{dietnetworks} A)**. This finding suggests that bacteria were targeted by less phages and that phage tropism was more specific. High-fat diets also exhibited decreased closeness centrality **(Figure \ref{dietnetworks} B)**, indicating that microbes were more distant from other microbes in the community. This would make genetic transfer and altered abundance less capable of impacting other bacteria and phages within the network.

@@

## Variation of Network Structure Across the Human Skin Landscape
Extensive work has illustrated differences in healthy human skin microbiome between anatomical sites, including bacteria, viruses, and fungi [@Grice:2009ee; @Findley:2013jf; @Hannigan:2015fz]. These communities vary by degree of skin moisture, oil, and environmental exposure. We hypothesized that like microbial composition and diversity, microbial network structure differed between anatomical sites. We tested this hypothesis by evaluating the changes in network structure between anatomical sites within our skin dataset.

@@

# Discussion
We developed and implemented a network-based analytical approach which we used to begin evaluating the basic properties of the human microbiome through bacteria and phage *relationships*, instead of membership or diversity. The goal of this study was to provide an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how microbiome networks differ in disease states. This goal of providing a baseline understanding of the human microbiome was similar to the goals of initial studies of bacteria and viral diversity across the human body, as well as other environments [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. Through the use of network theory, we leveraged extensive analytical opportunities which could be applied to understand complex ecological communities. We focused on utilizing metrics of connectivity to understand the extent to which communities of bacteria and phages interact (e.g. horizontal gene transfer, modulated bacterial gene expression, alterations in abundance). By pursuing this goal, we aimed to provide an initial understanding of human microbiome networks that will power future studies to use similar approaches in other setting such as disease states.

We reported that, just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were also conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggest that microbiome properties such as stability and the potential for horizontal gene transfer are personal, and potentially impacted by personal factors ranging from the body's immune system to environmental conditions such as climate and diet. The ability of environmental conditions to shape gut and skin microbiome network structure was supported by our finding that diet and skin location were associated with altered network structures.

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, we found evidence that high-fat diets were less connected than low-fat diets. This suggested that high-fat diets may lead to less stable communities with a decreased ability to influence each other. We supported this finding with the observation that obesity was associated with decreased network connectivity. Our results allow us to speculate that the food we eat may not only impact what microbes colonize our guts, but may also impact their interactions. Further work will be required to characterize these relationships with a larger cohort.

In addition to diet, we found evidence that skin microbiome network structure varied by skin environment. We observed that network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure, the exposed and intermittently exposed sites, were less connected and therefore were expected to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Thus body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes.

Our results represent a step toward understanding the microbiome through interspecies relationships. While these findings are informative, there are certainly caveats that should be noted. *First*, while our infection classification model is advantageous over existing models, we recognize that, like most classification models, there remains opportunity for improvement. For example, such a model is only as good as its training set, and future large-scale endeavors into infectious relationships (and the associated genomes) will provide more robust training and higher model accuracy. Just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will likely be integrated into this model to further improve model performance. *Second*, while informative, this work was done retrospectively and relied on published research from as many as seven years ago. These archived datasets were limited by the technology and costs of the time, meaning the datasets are poorly powered for the statistical analysis we strive for today. While we were able to present initial conclusions, follow-up studies will be required to validate our findings. Despite their limitations, these methods and results will be important for informing design and interpretation of future studies of the human microbiome.

>>>>

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Most previous work has relied on studying bacteria and phages in isolation, reducing them to two separate communities. Such approaches are not able to capture how these communities interact to maintain stability and share information (i.e. horizontal gene transfer). We developed and implemented a network-based analytical approach to provide an initial understanding of phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages infected which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate the diversity of phage-bacteria connectedness across the human body. We found that gut and skin network structures were person-specific, and were not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. Together this study characterized the diversity of virome-microbiome networks across the human body and illustrated how environmental factors may influence phage-bacteria interactive dynamics. This work provides an baseline for future studies to investigate the role of ecological networks in disease states.

**Word Count**: 238 / 250

@@

Viruses and bacteria are critical components to the human microbiome and play important roles in health and disease. Bacterial communities have been associated with diseases including a wide range of skin conditions [@Hannigan:2013im], acute and chronic wound healing conditions [@Hannigan:2014be; @Loesche:2016ev], and gastrointestinal diseases including inflammatory bowel disease [@He:2016ch; @Norman:2015kb], *Clostridium difficile* infections [@Seekatz:2016fz], and colorectal cancer [@Zackular:2014fba; @Baxter:2014hb]. Altered viromes (virus communities consisting primarily of bacteriophages in humans) have also been associated with various diseases and environmental perturbations including inflammatory bowel disease [@Norman:2015kb; @Manrique:2016dx], periodontal disease [@Ly:2014ew], spread of antibiotic resistance [@Modi:2013fia], and others [@Monaco:2016ita; @Hannigan:2015fz; @Minot:2011ez; @SantiagoRodriguez:2015gd; @Abeles:2015dy; @Abeles:2014kj]. These virus communities act in concert with bacteria as one single overall community [@Haerter:2014ii]. Individual phages within these communities are capable of lysing their bacterial hosts, modulating their functionality through processes such as horizontal gene transfer (i.e. transduction) [@Moon:2015fa, @Modi:2013fi, @Ogg:1981th, @Frost:2005dn], and altering bacterial host gene expression [@Lindell:2005gz, @Tyler:2013fl, @Hargreaves:2014ja].

Most previous work has relied on studying bacterial and phage communities in isolation, reducing them to two separate communities. In reality, bacteria and phage communities are dynamic and complex. They frequently share genetic information and work together to maintain stable ecosystems. Removal of members can disrupt or even collapse those ecosystems [@Haerter:2014ii; @Harcombe:2005fd; @Middelboe:2001fl; @Poisot:2011jc; @Thompson:2012ki; @Moebus:1981kp; @Flores:2013hc; @Poisot:2012fh; @Poisot071993; @Flores:2011bh; @Jover:2015ev]. The most common approach of studying communities in isolation has been unable to capture how these communities interact and influence one another, but relationship-based network approaches allow us to capture this information. We aimed to bridge this knowledge gap by leveraging machine learning and graph theory techniques to characterize human bacterial and phage communities by their relationships. We focused on characterizing the network diversity of the human skin and gut microbiomes to provide a foundation for further studies of disease network dynamics and to gain broader insights into human microbiome network diversity across different body sites.

First we investigated human microbiome phage and bacteria networks by utilizing three published microbiome datasets with paired virus and bacterial metagenomic sequence sets [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models by its inclusion of protein interaction data as well as both negative and positive interactions [@Edwards:2015iz, @Roux:2016cc].

Through this approach, we were able to go beyond contemporary isolated methods to provide a basic understanding of the community network dynamics associated with healthy human phages and bacteria. By building and utilizing a microbiome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. Through an improved understanding of the healthy state of network structures across the human body, we will empower future studies to begin investigating how these community structures change in disease and their overall impact on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in microbiome phage-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, including a study of the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. The viromes associated with these datasets were all subjected to virus-like particle (VLP) purification to eliminate other organismal DNA and allow for direct assessment of both the extracellular and actively replicating intracellular virome.

We utilized data that was published over the span of five years using different methods and technologies, and therefore yielded different sequence abundances **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. The bacterial and viral sequences from those publications were quality filtered and assembled into contigs. We generate clusters of related related bacterial and phage contigs by k-mer frequency and co-abundance using the CONCOCT algorithm. We interpreted these clusters as operational units of related bacteria and phage genomes that we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht]. The resulting contigs and OGUs demonstrated high sequence coverage and length **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**.

## Establishing a Model of Phage-Bacteria Interactions
We predicted which phage OGUs infected which bacteria using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. This training set contained 43 diverse bacterial species and 30 diverse phage strains, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because this was a DNA sequencing study, RNA phages were not considered **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (phage infects a bacterium) and negative relationships (phage does not infect a bacterium). This allowed us to validate the false positive and false negative rates associated with our candidate models, thereby building upon previous work that only considered positive relationship [@Edwards:2015iz].

Four phage and bacterial genomic features were used to predict infectious relationships between bacteria and phages: 1) genome nucleotide similarities, 2) gene amino acid sequence similarities, 3) CRISPR targeting of phages by bacterial CRISPR spacer sequences, and 4) similarity of protein families associated with experimentally identified protein-protein interactions [@Orchard:2014hq]. The resulting random forest model performed with an AUC of `r signif(genmodelper$ROC, digits = sigfig)`, a sensitivity of `r signif(genmodelper$Sens, digits = sigfig)`, and a specificity of `r signif(genmodelper$Spec, digits = sigfig)` **(Figure \ref{RocCurve} A)**. The most important predictor in the model was nucleotide similarity between genes, followed by nucleotide similarity of whole genomes **(Figure \ref{RocCurve} B)**. Protein family interactions were moderately important to the model, while CRISPRs were not informative important **(Figure \ref{RocCurve} B)**. Approximately one third of the relationships yielded no score and were therefore unknown, unable to be assigned an interaction prediction **(Figure \ref{RocCurve} C)**.

Next we used our random forest model to classify the relationships between bacteria and phage operational genomic units in the experimental datasets. The relationships predicted within the three published datasets were used to construct one master network, which was built to contain the three studies as sub-networks, which themselves each contained sub-networks for each sample **(Figure \ref{RocCurve} D)**. Metadata including study, sample ID, disease, and OGU abundance within the community were stored in the master network for downstream analysis **(Supplemental Figure \ref{NetworkDiagram})**. The master network was highly connected and contained `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Edges"], big.mark = ",")` infectious relationships among `r prettyNum(genfigurestats[c(genfigurestats$class %in% "Total"),"Vertices"], big.mark = ",")` nodes, `r prettyNum(nestats[c(nestats$cats %in% "PhageNodes"),"values"], big.mark = ",")` phages and `r prettyNum(nestats[c(nestats$cats %in% "BacteriaNodes"),"values"], big.mark = ",")` bacteria. Although the network was highly connected, not all relationships were present in all samples. As relationships were weighted by the relative abundances of their associated bacteria and phages, lowly abundant relationships could be present but not highly abundant. Like the master network, the skin network exhibited a diameter of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "SkinStudy"),"Diameter"], big.mark = ",")` (measure of graph size; the greatest number of traversed vertices required between two vertices) and included almost all of the master network nodes and edges **(Figure \ref{RocCurve} E - F)**. The phages and bacteria in the gut diet and twin sample sets were more sparsely related, each contained fewer than 150 vertices, fewer than 20,000 relationships, and diameters of `r prettyNum(genfigurestats[c(genfigurestats$class %in% "DietStudy"),"Diameter"], big.mark = ",")`, which indicated more sparsely related phages and bacteria **(Figure \ref{RocCurve} E - F)**.

## Individuality of Microbial Networks
Skin and gut community membership and diversity are highly personal, with people remaining more similar to themselves over time compared to other people [@Grice:2009ee; @Hannigan:2015fz; @Minot:2013ih]. We therefore hypothesized that this personal conservation extended to microbiome network structure. To answer this hypothesis, we calculated the degree of dissimilarity between each subject's network based on phage and bacteria abundance, as well as centrality. We quantified phage and bacteria centrality within each sample graph using the weighted eigenvector centrality metric. This metric defines central phages as those that are highly abundant and infect many distinct bacteria which themselves are abundant and infected by many other phages. Similarly, bacterial centrality was defined as those bacteria that were both abundant and connected to numerous phages that were themselves connected to many bacteria. We then calculated the similarity of community networks using the weighted eigenvector centrality of all nodes between all samples. Samples with similar network structures were interpreted as having similar capacities for maintaining stability and transmitting genetic material.

We used this network dissimilarity metric to test whether microbiome network structures were more similar within people than between people over time. We found that gut microbiome network structure clustered by person (ANOSIM p-value = `r signif(interstats[c(interstats$site %in% "DietAnosim"), "prob"], digits = sigfig)`, R = `r signif(interstats[c(interstats$site %in% "AS"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} A**). Network dissimilarity within people over the 8-10 day sampling period was less than the average dissimilarity between that person and others, although this was not statistically significant (p-value = `r signif(interstats[c(interstats$site %in% "Diet"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} B**). The lack of statistical confidence was likely due to the small gut microbiome sample size of this dataset. Although there was evidence for gut network conservation among individuals, we found no evidence for conservation of gut network structures within families. The gut network structures were not more similar between twins and their mothers (intrafamily) compared to other families of twins and mothers (inter-family) (p-value = `r signif(interstats[c(interstats$site %in% "Twins"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} C**).

<!-- Maybe add twin to twin data here? -->

Skin microbiome network structure was strongly conserved within individuals (p-value = `r signif(interstats[c(interstats$site %in% "Skin"), "prob"], digits = sigfig)`, **Figure \ref{intradiv} D**) and more adequately confirmed by our larger skin dataset. This distribution was similar when separated by anatomical sites, with most sites also being statistically significantly more conserved within individuals **(Supplemental Figure \ref{allskin})**.

## Role of Diet & Obesity in Gut Microbiome Conectivity

<!-- Could maybe move this section up before the previous section -->

Diet is a major environmental factor that influences resource availability and gut microbiome composition and diversity, including bacteria and phages [@Minot:2011ez; @Turnbaugh:2009hf; @David:2014cl]. Previous work in isolated culture-based systems has suggested that changes in nutrient availability are associated with altered phage - bacteria network structures [@Poisot:2011jc], although this has yet to be tested in humans. We hypothesized that a change in diet would be associated with a change in network structure in these human systems.

We evaluated the diet-associated differences in gut network structure by quantifying how central each sample's network was on average. We accomplished this by utilizing two common centrality metrics: degree centrality and closeness centrality. Degree centrality, the simplest centrality metric, was defined as the number of connections each phage made to bacteria, or each bacterium made to phages. We more completely quantified the network properties by supplementing degree centrality with measurements of closeness centrality. Closeness centrality is a metric of how close each phage or bacterium is to all of the other phages and bacteria in the network. A higher closeness centrality suggests that the effects of genetic information or altered abundance would be more impactful to all other microbes in the system. A network with higher average closeness centrality also indicates an overall greater degree of connections, which suggests a greater resilience against instability. Because these values are assigned to each phage and bacterium within each network, we calculated the average connectedness and corrected for the maximum potential degree of connectedness to obtain a single value for the connectedness of each graph.

We found that the gut microbiome network structures associated with high-fat diets were less connected than those of low-fat diets **(Figure \ref{dietnetworks} A-B)**. Tests for statistical differences were not performed due to the small sample size. High-fat diets exhibited less degree centrality **(Figure \ref{dietnetworks} A)**. This finding suggests that bacteria in high-fat environments were targeted by fewer phages and that phage tropism was more specific. High-fat diets also exhibited decreased closeness centrality **(Figure \ref{dietnetworks} B)**, indicating that bacteria and phages were more distant from other bacteria and phages in the community. This would make genetic transfer and altered abundance of a given phage or bacterium less capable of impacting other bacteria and phages within the network.

@@

## Association Between Environmental Stability and Network Structure Across the Human Skin Landscape
Extensive work has illustrated differences in healthy human skin microbiome between anatomical sites, including bacteria, viruses, and fungi [@Grice:2009ee; @Findley:2013jf; @Hannigan:2015fz]. These communities vary by degree of skin moisture, oil, and environmental exposure. We hypothesized that like microbial composition and diversity, microbial network structure would be specific to anatomical sites. We tested this hypothesis by evaluating the changes in network structure between anatomical sites within our skin dataset.

@@

# Discussion
Early, foundational microbiome work has provided a baseline understanding of the human microbiome by characterizing bacteria and viral diversity across the human body, as well as other environments [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. We similarily aimed to provide an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how microbiome networks differ in disease states. We developed and implemented a network-based analytical approach that we used to evaluate the basic properties of the human microbiome through bacteria and phage *relationships*, instead of membership or diversity. Through the use of network theory, we leveraged extensive analytical opportunities that could be applied to understand complex ecological communities. We utilized metrics of connectivity to model the extent to which communities of bacteria and phages interact (e.g. horizontal gene transfer, modulated bacterial gene expression, alterations in abundance). By pursuing this goal, we aimed to provide an initial understanding of human microbiome networks that will power future studies to use similar approaches in other settings, such as disease states.

We found that, just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were also conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggest that microbiome network properties, such as stability and the potential for horizontal gene transfer, are personal and may be impacted by personal factors ranging from the body's immune system to environmental conditions such as climate and diet. The ability of environmental conditions to shape gut and skin microbiome network structure was supported by our finding that diet and skin location were associated with altered network structures.

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, we found evidence that high-fat diets were less connected than low-fat diets. This suggested that high-fat diets may lead to less stable communities with a decreased ability for individuals to influence one other. We supported this finding with the observation that obesity was associated with decreased network connectivity. Together these findings suggest the food we eat may not only impact which microbes colonize our guts, but may also impact their interactions. Further work will be required to characterize these relationships with a larger cohort.

In addition to diet, we found that skin microbiome network structure varied by skin environment. Network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure, the exposed and intermittently exposed sites, were less connected and therefore were predicted to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Thus body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes.

Our results represent a step toward understanding the microbiome through interspecies relationships. While these findings are informative, there are certainly caveats that should be noted. *First*, while our infection classification model is advantageous over existing models, we recognize that, like most classification models, there remains opportunity for improvement. For example, such a model is only as good as its training set. Large scale experimental screens for phage and bacteria infectious interactions will provide more robust model training, and therefore improved model performance. Just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will be integrated into this model to further improve model performance.
